Reproducible development environments with rix

### 2023-12-14 paper.tex
rendered file

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, interior hidden, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Reproducible development environments with rix},
  pdfauthor={Bruno Rodrigues; Philipp Baumann},
  pdfkeywords={reproducibility, R, Nix},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% an abstract and keywords
\Abstract{In order to create an analysis that is easily reproducible, it
is not enough to write clean code and document it well. One must also
make sure to list all the dependencies of the analysis clearly and
ideally provide an easy way to install said dependencies. There are
several tools that can be used to list dependencies and to make them
easily installable by someone that wishes to reproduce a study, such as
{[}Docker{]}.\{proglang\}, a containerization solution. This paper will
present the Nix package manager, and an {[}R{]}.\{proglang\} package
called \pkg{rix} that lowers Nix's learning curve for users of the
{[}R{]}.\{proglang\} programming language.}

@@

\begin{verbatim}
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
\end{verbatim}

The \texttt{.Rprofile} file is a file that gets read by R when starting
a new session, which means that every call to the
\texttt{install.packages()} function will now install the packages from
this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be
to not only install the right packages used for the analysis, but also
the right version of \proglang{R}. Of course, it would be possible to
install the right version manually, but here too, there are tools that
simplify the process such as \citet{rlib2023}.

There are exceptions of course, a great example of a paper that provides
everything needed to reproduce its results is \citet{mcdermott2021}. The
author of this paper set up an accompagnying Github repository to the
paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
containing all the instructions to install the required software and
then run the analysis. If we take a closer look at this repository, we
will notice that several tools were used to capture the compatutational
environment and make it available to other researchers:

\begin{itemize}
\tightlist
\item
  The version of \proglang{R} was stated;
\item
  Packages and their versions were listed and saved into an
  \texttt{renv.lock} file;
\item
  A \texttt{Makefile} was used to run the whole analysis and compile the
  paper;
\item
  A \texttt{Dockerfile} was used to provide the complete computational
  environment, including the right version of \proglang{R} and run the
  whole analysis easily.
\end{itemize}

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, leftrule=.75mm, colback=white, rightrule=.15mm, left=2mm, opacityback=0, breakable, bottomrule=.15mm]

>>>>

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Reproducible development environments with rix},
  pdfauthor={Bruno Rodrigues; Philipp Baumann},
  pdfkeywords={reproducibility, R, Nix},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% an abstract and keywords
\Abstract{In order to create an analysis that is easily reproducible, it
is not enough to write clean code and document it well. One must also
make sure to list all the dependencies of the analysis clearly and
ideally provide an easy way to install said dependencies. There are
several tools that can be used to list dependencies and to make them
easily installable by someone that wishes to reproduce a study, such as
\proglang{Docker}, a containerization solution. This paper will present
the Nix package manager, and an \proglang{R} package called \pkg{rix}
that lowers Nix's learning curve for users of the \proglang{R}
programming language.}

@@

\begin{verbatim}
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
                               __linux__/jammy/2023-06-30"))
\end{verbatim}

The \texttt{.Rprofile} file gets read by R when starting a new session,
which means that every call to the \texttt{install.packages()} function
will now install the packages from this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be
to not only install the right packages used for the analysis, but also
the right version of \proglang{R}. Of course, it would be possible to
install the right version manually, but here too, there are tools that
simplify the process such as \proglang{rig} by the R infrastructure team
\citeyearpar{rlib2023}. One could thus install the right version of
\proglang{R} using \proglang{rig} allows to easily install different
versions of \proglang{R}, one could use to install the right version to
reproduce an analysis, and then use one of the listed packages above to
install the right library of of \proglang{R} packages. However, this
involves many manual steps and is thus error prone. It is also
time-consuming to do so, especially if one wants to reproduce many
studies.

The final step towards the gold standard of reproducibility is to
package the right version of \proglang{R} and \proglang{R} packages
inside a \proglang{Docker} image. \proglang{Docker} is a
containerisation tool: using \proglang{Docker} it is possible to package
some \emph{data product} with its dependencies into an image. From this
image, containers can be run to reproduce the data product with its
exact dependencies. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that comes with
many software dependencies. \emph{Dockerizing} an analysis consists in
first building an image: in the build step, the dependencies of the
analysis have to be installed and this can be achieved using the tools
mentioned above. The scripts to run the analysis and the data are also
added to the image at build time. Instances of that image, called
containers, can then be executed, which give access to the exact
development environment originally used for the study. It is also
possible to script the building of the analysis itself, such that in one
single call to the \texttt{docker\ run} command, a complete analysis is
reproduced.

The Rocker project initiated by \citet{boettiger2017} provides many
pre-built \proglang{Docker} to the \proglang{R} community of users.
These images can be used as bases to build other images containing
statistical analyses more easily than starting from a bare-bones image.

An optional step towards the gold standard is to use a build automation
tool such as \proglang{Make} to run the whole analysis when the
container is executed. Build automation tools make it easier to run
arbitrary code in a series of well-defined steps. \proglang{R}
programmers can use the \pkg{targets} by \citet{landau2021} as a build
automation tool.

\citet{mcdermott2021} is an example of a scientific study that reached
the gold standard of reproducibility. The author of this paper set up an
accompagnying Github repository to the paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
containing all the instructions to install the required software and
then run the analysis. If we take a closer look at this repository, we
will notice that many of the tools previously mentioned were used to
capture the compatutational environment and make it available to other
researchers:

\begin{itemize}
\tightlist
\item
  Packages and their versions were listed and saved into an
  \texttt{renv.lock} file;
\item
  A \texttt{Makefile} was used to run the whole analysis and compile the
  paper;
\item
  A \texttt{Dockerfile} was used to provide the complete computational
  environment, including the right version of \proglang{R} and run the
  whole analysis easily.
\end{itemize}

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, breakable, left=2mm, colback=white, arc=.35mm, rightrule=.15mm, toprule=.15mm, bottomrule=.15mm]
